---
title: "The relationship between Gross Domestic Product per capita and Life Expectancy."
author: "Emma Rand"
engine: knitr
format:
  html
  # pdf
execute:
  echo: false
  include: true
  error: false
  message: false
  warning: false
bibliography: references.bib
params:
  year: 2002
---

```{r}
#| label: load-packages
library(tidyverse)
```

# Introduction

There is a relationship between wealth and health [@pollack_should_2007]. We
will explore this relationship using the relationship between [Gross Domestic
Product (GDP)](https://en.wikipedia.org/wiki/Gross_domestic_product) *per
capita* and Life Expectancy across countries and over time.

```{r}
#| label: import-data
file <- "data-raw/gapminder.tsv"
gapminder <- read_tsv(file)
```

```{r}
#| label: extract-summary
# extract values for inline reporting
n_years  <- length(unique(gapminder$year))
min_year <- min(gapminder$year)
max_year <- min(gapminder$year)
n_counties <- length(unique(gapminder$country))
```

# Methods

We took per capita GDP and life expectancy data from `r n_years` years (from
`r min_year` to `r max_year`) for each of `r n_counties` countries. Analysis was
carried out with R [@R-core] and tidyverse [@tidyverse] packages. This report is
for the year `r params$year`.

# Results

Data about about the five most populated countries in is shown in
@tbl-year-top-pop.

```{r}
#| label: year-top-pop
year_top_pop <- gapminder |> 
  filter(year == params$year) |> 
  arrange(desc(pop)) |> 
  slice_head(n = 5) 
```

::: {#tbl-year-top-pop}
```{r}
year_top_pop |> 
  knitr::kable()
```

Five most populated countries in `r params$year`.
:::

Data about about the five most wealthy countries in is shown in
@tbl-year-top-gdp.

```{r}
#| label: year-top-gdp
year_top_gdp <- gapminder |> 
  filter(year == params$year) |> 
  arrange(desc(gdpPercap)) |> 
  slice_head(n = 5) 
```

::: {#tbl-year-top-gdp}
```{r}
year_top_gdp |> 
  knitr::kable()
```

Five most wealthy countries in `r params$year`.
:::

The relationship between Gross Domestic Product (GDP) per capita and Life
Expectancy across countries in `r params$year` is in @fig-gdp-life.

```{r}
#| label: subset-year

# Filter the dataset to the report year
gapminder_year <- gapminder |> 
  filter(year == params$year)
```

::: {#fig-gdp-life}
```{r}
# Create a bubble chart
ggplot(gapminder_year, aes(x = gdpPercap, 
                               y = lifeExp, 
                               color = continent, 
                               size = pop/10^6)) +
  geom_point() +
  scale_size_continuous(name = "Population\n(millions)") +
  scale_color_discrete(name = NULL) +
  scale_x_continuous(name = "Gross Domestic Product per capita") +
  scale_y_continuous(name = "Life Expectancy (years)") +
  theme_classic()

```

**The relationship between Gross Domestic Product (GDP) *per capita* and Life
Expectancy across countries in `r params$year`.** Each point represents a
country, plotted by gross domestic product (GDP) *per capita* (international
dollars, x-axis) against life expectancy at birth in years (y-axis). Point
colour indicates continental region and point size is proportional to population
size (millions). Data are from Gapminder and were analysed with R and tidyverse
packages.
:::

# Discussion

# References
