---
title: "The relationship between Gross Domestic Product per capita and Life Expectancy."
author: "Emma Rand"
engine: knitr
format:
  html
execute:
  echo: false
  include: true
  error: false
  message: false
  warning: false
---

```{r}
#| label: load-packages
library(tidyverse)
```

# Introduction

There is a relationship between wealth and health. We will explore this
relationship using the relationship between [Gross Domestic Product
(GDP)](https://en.wikipedia.org/wiki/Gross_domestic_product) *per capita* and
Life Expectancy across countries and over time.

```{r}
#| label: import-data
file <- "data-raw/gapminder.tsv"
gapminder <- read_tsv(file)
```

```{r}
#| label: extract-summary
# extract values for inline reporting
n_years <- length(unique(gapminder$year))
min_year <- min(gapminder$year)
max_year <- max(gapminder$year)
n_counties <- length(unique(gapminder$country))
```

# Methods

We took per capita GDP and life expectancy data from `r n_years` years (from
`r min_year` to `r max_year`) for each of `r n_counties` countries. Analysis was
carried out with R and tidyverse packages.

# Results

Data about about the five most populated countries in is shown in
@tbl-max-year-top-pop.

```{r}
#| label: max-year-top-pop
max_year_top_pop <- gapminder |> 
  filter(year == max_year) |> 
  arrange(desc(pop)) |> 
  slice_head(n = 5) 
```

::: {#tbl-max-year-top-gdp}
```{r}
#| label: tbl-max-year-top-pop
max_year_top_pop |> 
  knitr::kable()
```

Five most populated countries in `r max_year`.
:::

Data about about the five most wealthy countries in is shown in
@tbl-max-year-top-gdp.

```{r}
#| label: max-year-top-gdp
max_year_top_gdp <- gapminder |> 
  filter(year == max_year) |> 
  arrange(desc(gdpPercap)) |> 
  slice_head(n = 5) 
```

::: {#tbl-max-year-top-gdp}
```{r}
max_year_top_gdp |> 
  knitr::kable()
```

Five most wealthy countries in `r max_year`.
:::

# Discussion

# References
